The Effects of the NBA COVID Bubble on the NBA Playoffs: A Case Study for Home-Court Advantage

### 2022-02-04 Manuscript/manuscript.tex
Update manuscript.tex
Updated Tex for submission file

\documentclass[12pt, letterpaper, titlepage]{article}

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage {hyperref}
\hypersetup {colorlinks = true, linkcolor = blue, citecolor=blue, urlcolor = blue}
\usepackage[]{natbib}
\usepackage{enumitem}
\usepackage{setspace}

\usepackage[pagewise]{lineno}
\linenumbers*[1]
% %% patches to make lineno work better with amsmath
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
 {\linenomath\csname old#1\endcsname}%
 {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%

\AtBeginDocument{%
 \patchBothAmsMathEnvironmentsForLineno{equation}%
 \patchBothAmsMathEnvironmentsForLineno{align}%
 \patchBothAmsMathEnvironmentsForLineno{flalign}%
 \patchBothAmsMathEnvironmentsForLineno{alignat}%
 \patchBothAmsMathEnvironmentsForLineno{gather}%
 \patchBothAmsMathEnvironmentsForLineno{multline}%
}

\title{The Effects of the NBA COVID Bubble on the NBA Playoffs: A Case Study for Home-Court Advantage}

% \author{Michael Price and Jun Yan\\
% \href{mailto:michael.price@uconn.edu}{\nolinkurl{michael.price@uconn.edu}}\\
% Department of Statistics, University of Connecticut}
\author{annonymous}
\date{December 1, 2020}

\begin{document}
\maketitle

\doublespace

\begin{abstract}
The 2020 NBA playoffs were played inside of a bubble in Disney World because of
the COVID-19 pandemic. This meant that there were no fans in attendance,
games played on neutral courts and no traveling for teams, which in theory
removes home-court advantage from the games. This setting has attracted much
discussion as analysts and fans debated the possible effects it may have on the
outcome of games. Home-court advantage has historically played an influential
role in NBA playoff series outcomes. The 2020 playoff provided a unique
opportunity to study the effects of the bubble and home-court
advantage by comparing the 2020 season with the seasons in the past.
While many factors contribute to the outcome of games, points scored
is the deciding factor of who wins games, so scoring is the primary focus
of this study. The specific measures of interest are team scoring totals and
team shooting percentage on two-pointers, three-pointers, and free throws.
Comparing these measures for home teams and away teams in 2020 vs.~2017-2019
shows that the 2020 playoffs favored away teams more than usual, particularly
with two point shooting and total scoring.
\end{abstract}

\hypertarget{sec:intro}{%
\section{Introduction}\label{sec:intro}}

Home-court advantage is often discussed in sports circles as a contributing
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge from playing at their home-court, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a
bubble due to the COVID-19 pandemic
brought a great deal of concern for fans, teams, journalists, and others.
For example, \citet{Aschburner} discusses about anticipated effects, sharing
concerns from former players, coaches and other experts about
the potential effects of removing home-court advantage. Aschburner notes that the
NBA did make attempts to recreate the effects by putting the ``home'' team logo on
the court and allowing the ``home'' team to play crowd noise and music, but most
people doubted these small attempts would recreate a true playoff atmosphere.
During the 2020 NBA playoffs, home teams only won about 48.2\% of the games. This
is lower than normal, which Aschburner claims usually floats around 60\%. This
shift in the home team winning percentage surely indicates the opportunity for
thorough investigation.

@@

Our study is quite different from earlier NBA home-court advantage studies.
By using the neutral site games of 2020 we will get to compare home and away
performance to a control. Typically, studies just compare home vs away
performance. These studies do not separate the effects of home-court advantage
into the specific effect on the home team and the specific effect on the away
team. They show that home teams outperform away teams, but not if this is a
result of home teams overperforming or away teams underperforming because of
home-court advantage. Some of these studies are reviewed in greater detail in
Section\textasciitilde\ref{sec:litrev}.

We will compare home team performance in 2020 at a neutral site with
no fans vs.~2017-19 playoffs with fans. Likewise, away team performance in 2020
at a neutral site with no fans vs.~2017-19 playoffs with fans. By comparing home
teams in 2020 to home teams in 2017-19 and away teams in 2020 to away teams in
2017-19, we add a new perspective to the field of research. This will allow for
a more accurate understanding of the effects of home-court advantage on home and
away teams in the NBA. We will not only see that home-court advantage helps home
teams outperform away teams, but also separate the
effects of home-court advantage on home teams and away teams performance individually.

Nine hypotheses were tested to understand the differences in 2020 vs.~earlier
years. First, whether or not the difference between home win percentage in 2020
and 2017-19 is zero. This difference is found to be statistically
significant from zero. Then we assess for differences in home
scoring in 2020 vs 2017-2019. Similarly, we can do the same test, but for
differences in away scoring in 2020 vs 2017-2019. Also, differences in team
shooting (for-two pointers, three-pointers, and free throws) from 2020 vs 2017-2019
for both home and away teams. The results from these tests bring a new perspective
to understanding of how home-court advantage impacts games by altering the performance
of the home and away teams.

\hypertarget{sec:litrev}{%
\section{Literature Review}\label{sec:litrev}}

There is a voluminous literature on the effects of home-court advantage.
Many NBA home-court advantage studies analyze the effects by studying shooting
percentages. \citet{Kotecki} reported significant home-court advantage using performance-based
statistics, specifically field goal percentage, free throw percentage, and
points by comparing home performance vs.~away performance in games.
All of which he showed significantly indicates that
home-court helps teams play better. \citet{Cao} studied the effects
of pressure on performance in the NBA. Using throws as their measure of
interest, they tested whether home
fans could distract and put pressure on opposing players to make free
throws. However, they found insignificant
evidence that home status has a substantial impact on missing from the free throw
line. \citet{Harris} used two point shots, three-point shots and free throws as measures
of interest to study home-court advantage. Two point shots were found to be
the strongest predictor of home-court advantage. They suggested
that home teams should try to shoot more two point shots and force
their opponent to take more two point shot attempts. This strategy will maximize
the benefits of home-court advantage and give them the best chance to win.

Some studies focus less on shooting and more on scoring differences and other metrics.
For example, \citet{Greer} focused on the influence of spectator booing on
home-court advantage in basketball. The three methods of performance used
in this study were scoring, violations, and turnovers. This study was conducted
using the men's basketball programs at two large universities. The study finds that
social support, like booing, is an important contributor to home-court advantage.
Greer explains, whether the influence is greater on visiting team performance or
referee calls is less clear. However, the data does seem to lean slightly in
favor of affecting visiting team performance. \citet{Harville} studied the
effect of home-court advantage using the 1991-1992 college basketball season.
Unlike the NBA, it is not uncommon to have a few games played at neutral sites
during the college basketball season. This allowed them to construct two samples,
one of home teams and one of neutral teams. They formulate their study in a
regression predicting the expected difference in score for home teams.
They set up their study to find if the home teams won games by more points when
they had home-court advantage vs.~when playing at neutral court. This study
concluded with evidence supporting home-court advantage.

There are also surveys on the factors contributing to home-court advantage.
\citet{Carron1992} gave four main game location factors for home and away
teams, namely, the crowd factor, which is the impact of fans cheering; learning
factors, which is an advantage from home teams from playing at
a familiar venue; travel factors, the idea that away teams may face
fatigue and jet lag from traveling; and, rule factors, which says that home teams
may benefit from some advantages in rules and officiating. They acknowledge that
these factors would all be removed if games were played at a neutral site even
if one team was designated as ``home team''. This study was reviewed a decade later
by \citet{Carron2005}. The 2005 review goes over the new findings from studies
about the significance of these four game location factors. Since 1992 they found
that results on these four factors are mixed. However, there is some evidence
supporting crowd and travel factors impact games in the NBA. There is less
evidence suggesting learning and rule factors impact the NBA. One interesting
finding cited by \citet{Carron2005} is that in the absence of crowds result
in overall performance increases.

\hypertarget{sec:data}{%
\section{Data}\label{sec:data}}

Data were collected from the official NBA website. The main variables of interest
are whether or not the home team won, scoring totals for home and away teams, and
shooting percentages for home and away teams on two-pointers, three-pointers, and
free throws. These variables were very popular and frequently used in the related
literature discussed earlier. The data was collected on a game by game basis, this gave us two observations
for each variable per game played, one observation for each team(home and away).
There were 83 games played in the 2020 playoffs, giving 83 observations
for each variable in 2020 for both the home and away teams (166 observations total).
Likewise, there were 243 games played over 2017-2019, giving 243 observations of each variable
for both home and away teams over 2017-2020 (486 observations total).
While many other measures could be used for measuring
the outcome of the game and team performance, scoring seemed to be the most
important. The winner of a game is determined by who scores more points. There can only be
one winner and one loser making the outcome a binary variable, with
one indicating a win and zero indicating a loss.

Home-court advantage is the basic idea that the home team is more likely to win.
So laying a foundation of typical home-court advantage is crucial. Before
focusing on the 2017 to 2020 playoffs we can take a quick look at home team win
percentages since 2010. Notice in Figure \ref{fig:Fig1}, the 10 years before 2020, the home team
winning percentage ranged from around 0.56 to 0.7 and never dipped below 0.5. The
2020 bubble broke this historic pattern dipping down below 0.5. Foreshadowing,
the confirmation of the expectation that the effect of home-court advantage was removed
in the 2020 playoffs.

\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{Fig1}
  \caption{Winning percentage of NBA home teams in the playoffs since
    2010, the green line denotes .500.}
  \label{fig:Fig1}
\end{figure}

Moving on to the main focus of the study, comparing 2020 to 2017-2019.
Figure \ref{fig:Fig2} shows the histograms of the home (green) and
away scoring (red) for 2020 vs.~2017-2019. All histograms are
fairly bell shaped, which is important for statistical tests
designed for normally distribute data. There appears to be little
difference between the 2020 and 2017-2019 for home scoring, but for
away scoring, a noticeable shift to the right in 2020 is observed
compared with that in 2017-2019.

\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2}
  \caption{Histograms of home (blue) and away (red) scoring
    for 2020 (bottom) and 2017-2019 (top).}
  \label{fig:Fig2}
\end{figure}

Our second target of inference is shooting percentage for home and away teams.
Figure \ref{fig:Fig3} shows home shooting for two-pointers, three-pointers and
free throws for 2020 (top) vs.~2017-19 (bottom). The histograms appear to be fairly
similarly distributed between 2020 and 2017-19. Likewise, Figure \ref{fig:Fig4},
shows the same percentages except for away teams. It appears that the
two-point shooting percentage for away teams has a small shift to the
right in 2020 relative to that in 2017-2019.

\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{Fig3}
  \caption{Histograms of home shooting percentages for two-
    pointers, three-pointers and free throws for 2020 (top) vs.~2017-19
    (bottom).}
  \label{fig:Fig3}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{Fig4}
  \caption{Histograms of away shooting percentages for two-pointers,
    three-pointers and free throws for 2020 (top) vs.~2017-19
    (bottom).}
  \label{fig:Fig4}  
\end{figure}

\hypertarget{sec:methods}{%
\section{Methods}\label{sec:methods}}

The 2020 bubble provides a new and exciting opportunity to study home-court
advantage for the NBA. Unlike college basketball, aside from a few
exhibition/preseason games, the NBA always has a home and away team. So, for
the first time in NBA history the bubble allows NBA home and away performance
to be compared vs a control/neutral field. The NBA bubble, as a neutral court,
removed all 4 possible game location factors impacting home-court advantage
hypothesized by \citet{Carron1992}. The NBA bubble featured 8 seeding games then a
standard playoff format. The focus of this study was the play during the playoff
games since it followed the standard playoff format and can easily be compared
back to other playoffs. For this study, the 2020 playoffs were compared against
the three previous playoffs collectively. To control
for the changing play style of the NBA, we limit the study to 2020 vs 2017-2019
for the faster pace play and more common use of the three-point shot
in modern basketball. If we used data from say 10 years ago, or earlier, observed
differences may not be from effects of the NBA bubble, but rather from the effects
of drastic changes in the style of play between the seasons. However, basketball
evolves slow enough that we can reasonably assume 2017-2019 are at least very
close in pace and playing style to 2020.

Comparisons between 2020 and 2017-19 home and away teams were made on home team
winning percentage, total team scoring and two-point, three-point and free throw
shooting. Comparing the differences in these metrics for home and away teams in
2020 vs previous years will provide valuable insights to the understanding of
home-court advantage. We can see how going on the road may negatively impact
away performance and how playing at home may positively impact home performance.
If there are differences in scoring for home or away teams the differences can
be used to show how home-court advantage affects overall performance of home
and away teams. While testing for differences in shooting will provide added
context for how home-court advantage specifically affects performance.
Shooting percentages are not the only possible metrics affected by home-court
advantage, but they are the most obvious and certainly an important one.

@@

We also conducted nonparametric tests that are distribution free to
confirm the results from the \(z\)-test. For question\textasciitilde1, we used
Fisher's exact test for a contigency table which summarizes the wins
and losses of the home team in the 83 games in 2020 and the 243 games
in 2017-2019. For all other eight questions, the data are the scores
or shooting percentages from the 84 games in 2020 and the 243 games in
2017-2019. We used Wilcoxan's rank-sum test.

@@

\hypertarget{sec:results}{%
\section{Results}\label{sec:results}}

\begin{table}
  \caption{The results from the 9 tests.}
  \label{tab:table}
\centering
\begin{tabular}[t]{lcccccc}
  & 2020 & 2017-19 & \multicolumn{2}{c}{P-value} & \multicolumn{2}{c}{Adjusted P-value}\\
  &          &                & Z-test & Wilcoxan & Z-test & Wilcoxan\\
Home Win & 0.482 & 0.613 & 0.0497* & 0.0400* & .4473 & 0.3598\\
Home Scoring & 1.101 & 1.081 & 0.2321 & 0.2985 & 1 & 1\\
Away Scoring & 1.091 & 1.040 & 0.0008* & 0.0004* & 0.0075* & 0.0035*\\
Home 2P & 0.523 & 0.515 & 0.4335 & 0.5719 & 1 & 1\\
Home 3P & 0.363 & 0.357 & 0.5733 & 0.8852 & 1 & 1\\
Home FT & 0.793 & 0.774 & 0.0692 & 0.0496* & 0.6228 & 0.4464\\
Away 2P & 0.536 & 0.504 & 0.0003* & 0.0003* & 0.0030 & 0.0023\\
Away 3P & 0.357 & 0.346 & 0.3256 & 0.3081 & 1 & 1\\
Away FT & 0.783 & 0.777 & 0.6601 & 0.8370 & 1 & 1\\
\end{tabular}
\end{table}

Starting from the top of the
Table \ref{tab:table} summarizes p-values of the nine hypotheses for both
\(z\)-tests and Wilcoxon tests. The p-values are all fairly similar for both
tests giving strong confidence in conclusions drawn. Also reported are
the point estimates of the two samples in each comparison.

>>>>

% Created by Bonita Graham
% Last update: December 2019 By Kestutis Bendinskas

% Authors: 
% Please do not make changes to the preamble until after the solid line of %s.

\documentclass[10pt]{article}
\usepackage[explicit]{titlesec}
\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}
\usepackage{hyphenat}
\usepackage{ragged2e}
\RaggedRight

% These commands change the font. If you do not have Garamond on your computer, you will need to install it.
\usepackage{garamondx}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amsthm}
\usepackage{graphicx}

% This adjusts the underline to be in keeping with word processors.
\usepackage{soul}
\setul{.6pt}{.4pt}

% The following sets margins to 1 in. on top and bottom and .75 in on left and right, and remove page numbers.
\usepackage{geometry}
\geometry{vmargin={1in,1in}, hmargin={.75in, .75in}}
\usepackage{fancyhdr}
\pagestyle{fancy}
\pagenumbering{gobble}
\renewcommand{\headrulewidth}{0.0pt}
\renewcommand{\footrulewidth}{0.0pt}

% These Commands create the label style for tables, figures and equations.
\usepackage[labelfont={footnotesize,bf} , textfont=footnotesize]{caption}
\captionsetup{labelformat=simple, labelsep=period}
\newcommand\num{\addtocounter{equation}{1}\tag{\theequation}}
\renewcommand{\theequation}{\arabic{equation}}
\makeatletter
\renewcommand\tagform@[1]{\maketag@@@ {\ignorespaces {\footnotesize{\textbf{Equation}}} #1.\unskip \@@italiccorr }}
\makeatother
\setlength{\intextsep}{10pt}
\setlength{\abovecaptionskip}{2pt}
\setlength{\belowcaptionskip}{-10pt}

\renewcommand{\textfraction}{0.10}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\floatpagefraction}{0.90}

% These commands set the paragraph and line spacing
\titleformat{\section}
  {\normalfont}{\thesection}{1em}{\MakeUppercase{\textbf{#1}}}
\titlespacing\section{0pt}{0pt}{-10pt}
\titleformat{\subsection}
  {\normalfont}{\thesubsection}{1em}{\textit{#1}}
\titlespacing\subsection{0pt}{0pt}{-8pt}
\renewcommand{\baselinestretch}{1.15}

% This designs the title display style for the maketitle command
\makeatletter
\newcommand\sixteen{\@setfontsize\sixteen{17pt}{6}}
\renewcommand{\maketitle}{\bgroup\setlength{\parindent}{0pt}
\begin{flushleft}
\sixteen\bfseries \@title
\medskip
\end{flushleft}
\textit{\@author}
\egroup}
\makeatother

% This styles the bibliography and citations.
%\usepackage[biblabel]{cite}
\usepackage[sort&compress]{natbib}
\setlength\bibindent{2em}
\makeatletter
\renewcommand\@biblabel[1]{\textbf{#1.}\hfill}
\makeatother
\renewcommand{\citenumfont}[1]{\textbf{#1}}
\bibpunct{}{}{,~}{s}{,}{,}
\setlength{\bibsep}{0pt plus 0.3ex}

\usepackage{float}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Authors: Add additional packages and new commands here.  
% Limit your use of new commands and special formatting.

% Place your title below. Use Title Capitalization.
\title{The Effects of the NBA COVID Bubble on the NBA Playoffs: A Case Study for Home-Court Advantage}

% Add author information below. Communicating author is indicated by an asterisk, the affiliation is shown by superscripted lower case letter if several affiliations need to be noted.
\author{
Michael Price*$^{a}$ \\ \medskip 
$^{a}$Department of Statistics, University of Connecticut, Storrs, CT  \\  \medskip 
\\ \medskip 
Students: michael.price@uconn.edu* \\
Mentor: jun.yan@uconn.edu 
}

\pagestyle{empty}
\begin{document}

% Makes the title and author information appear.
\vspace*{.01 in}
\maketitle
\vspace{.12 in}

% Abstracts are required.
\section*{abstract}
The 2020 NBA playoffs were played inside of a bubble at Disney World because of
the COVID-19 pandemic. This meant that there were no fans in attendance,
games were played on neutral courts and no traveling for teams. In theory, these conditions should
remove home-court advantage from the games. This setting generated
discussion and concern, as analysts and fans debated the possible effects it may have on the
outcome of games. Home-court advantage has historically played an influential
role in NBA playoff series outcomes. The 2020 playoffs provided a unique
opportunity to study the effects of the bubble and home-court
advantage by comparing the 2020 season with the seasons in the past.
While many factors contribute to the outcome of games, points scored
is the deciding factor of who wins. Thus, scoring is the primary focus
of this study. The specific measures of interest are team scoring totals and
team shooting percentage on two-pointers, three-pointers, and free throws.
Comparing these measures for home teams and away teams in 2020 vs.~2017-2019
shows that the 2020 playoffs favored away teams more than usual, particularly
with two-point shooting and total scoring.

% Keywords are required.
\section*{keywords} 
NBA; NBA Covid; NBA Bubble; HomeCourt Advantage;

\vspace{.12 in}

% Start the main part of the manuscript here.
% Comment out section headings if inappropriate to your discipline.
% If you add additional section or subsection headings, use an asterisk * to avoid numbering.

\section{Introduction}

Home-court advantage is often discussed in sports circles as a contributing
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge while playing at their homecourt, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a
bubble due to the COVID-19 pandemic
brought a great deal of concern for fans, teams, journalists, and others. The bubble
environment would not be able to replicate many of the factors that normal home-court
advantage relies on. Among the biggest is the effect from the crowd. The home crowd has
the ability to energize home teams, help control momentum in favor of the home team and 
create a chaotic and stressful environment for away teams to play in. There is no way
to replicate these effects in the bubble conditions. Also, there are geographic factors
that are lost, like altitude and time-zone effects. There's evidence across all sports
that teams in high altitude regions can rely on away teams coming in and struggling with
the lower oxygen levels at a higher altitude, which the home teams are accustomed to playing
with \citet{Lopez}. Examples of high altitude teams in the NBA would be the Utah Jazz
and Denver Nuggets, both of whom participated in the 2020 NBA bubble. Also, referees 
are known to be subject to favoring home teams. This is due in large part to pressure
from fans, which is no longer in play in the bubble. Refs may still have biases, but 
there's nothing to sway that bias towards the home team like normal. Any positive
effects of the home team playing in the arena they're most comfortable in and
living in the comfort of their own home are also lost. Travel is also normally 
discussed as a factor hurting away teams, but in the context of the playoffs
being played in the bubble, losing this may not be significant. In the playoffs,
many times (aside from the first game of the series when the away team travels), 
both teams often travel at the same time, since they're both going back and forth
between the two cities to complete the series.

 \citet{Aschburner} discusses the anticipated effects, sharing
concerns from former players, coaches and other experts about
the potential effects of removing home-court advantage. Aschburner notes that the
NBA did make attempts to recreate the effects by putting the ``home'' team logo on
the court and allowing the ``home'' team to play crowd noise and music, but most
people doubted these small attempts would recreate a true playoff atmosphere.
During the 2020 NBA playoffs, home teams only won about 48.2\% of the games. This
is lower than normal, which Aschburner claims usually floats around 60\%. This
shift in the home team winning percentage surely indicates the opportunity for
thorough investigation.

@@

Our study is quite different from earlier NBA home-court advantage studies.
By using the neutral site games of 2020, we will get to compare home and away
performance to a control. Typically, studies just compare home vs away
performance. These studies do not separate the effects of home-court advantage
into the specific effect on the home team and the specific effect on the away
team. They show that home teams outperform away teams, but not if this is a
result of home teams overperforming or away teams underperforming because of
home-court advantage. Some of these studies are reviewed in greater detail in Section 2.

We will compare home team performance in 2020 at a neutral site with
no fans vs.~2017-19 playoffs with fans. Likewise, away team performance in 2020
at a neutral site with no fans vs.~2017-19 playoffs with fans. By comparing home
teams in 2020 to home teams in 2017-19 and away teams in 2020 to away teams in
2017-19, we add a new perspective to the field of research. This will allow for
a more accurate understanding of the effects of home-court advantage on home and
away teams in the NBA. We will not only see that home-court advantage helps home
teams outperform away teams, but also separate the
effects of home-court advantage on home teams' and away teams' performance individually.

Nine hypotheses were tested to understand the differences in 2020 vs.~earlier
years. First, whether or not the difference between home win percentage in 2020
and 2017-19 is zero. This difference is found to be statistically
significant from zero. Then we assess for differences in home
scoring in 2020 vs 2017-2019. Similarly, we can do the same test, but for
differences in away scoring in 2020 vs 2017-2019. Also, differences in team
shooting (for two-pointers, three-pointers, and free throws) from 2020 vs 2017-2019
for both home and away teams. The results from these tests bring a new perspective
to the understanding of how home-court advantage impacts games by altering the performance
of the home and away teams.

\section{Literature Review}
There is voluminous literature on the effects of home-court advantage.
Many NBA home-court advantage studies analyze the effects by studying shooting
percentages. \citet{Kotecki} reported significant evidence of home-court
advantage by comparing field goal
percentage, free throw percentage and points scored in home vs.~away teams.
He found all of these measures showed evidence that
home-court advantage helps home teams play better. \citet{Cao} studied the effects
of pressure on performance in the NBA. Using free throws as their measure of interest, 
they tested whether home
fans could distract and put pressure on opposing players to make free
throws. However, they did not find significant evidence
that home status has a substantial impact on missing from the free throw line. 
\citet{Harris} used two-point shots, three-point shots and free
throws as measures of interest to study home-court advantage. Two-point
shots were found to be the strongest predictor of home-court advantage. They
suggested that home teams should try to shoot more two-point shots and force
their opponent to take more two-point shot attempts. This help home teams have a
greater control of the game play and help maximize the benefits 
home-court advantage.

Some studies focus less on shooting and more on other metrics.
For example, \citet{Greer} focused on the influence of spectator booing on
home-court advantage in basketball. The three methods of performance used
in this study were scoring, violations, and turnovers. This study was conducted
using the men's basketball programs at two large universities. The study finds that
social support, like booing, is an important contributor to home-court advantage.
Greer explains, whether the influence is greater on visiting team performance or
referee calls is less clear. However, the data does seem to lean slightly in
favor of affecting visiting team performance. \citet{Harville} studied the
effect of home-court advantage using the 1991-1992 college basketball season.
Unlike the NBA, it is not uncommon to have a few games played at neutral sites
during the college basketball season. This allowed them to construct two samples,
one of home teams and one of neutral teams. They formulated their study in a
regression predicting the expected difference in score for home teams.
They set up their study to find if the home teams won games by more points when
they had home-court advantage vs.~when playing at neutral court. This study
concluded with evidence supporting home-court advantage.

There are also surveys on the factors contributing to home-court advantage.
\citet{Carron1992} gave four main game location factors for home and away
teams, namely, the crowd factor, which is the impact of fans cheering; learning
factors, which is an advantage from home teams from playing at
a familiar venue; travel factors, the idea that away teams may face
fatigue and jet lag from traveling; and, rule factors, which says that home teams
may benefit from some advantages in rules and officiating. They acknowledge that
these factors would all be removed if games were played at a neutral site, even
if one team was designated as ``home team''. This study was reviewed a decade later
by \citet{Carron2005}. The 2005 review goes over the new findings from studies
about the significance of these four game location factors. Since 1992, they have found
that results on these four factors are mixed. However, there is some evidence
supporting crowd and travel factors impact games across all major sports. There is less
evidence suggesting learning and rule factors impact across the various collegiate 
and professional sports. The NBA is not a league which has rules that
may favor the home team, like batting last in the MLB, but these rule
factors also account for referee bias which may impact the NBA.  One
interesting
finding cited by \citet{Carron2005} is that the absence of crowds results
in overall performance increases. Lastly, as discussed in the introduction, \citet{Lopez}
find evidence that geographic factors like altitude may influence and strengthen home-court
advantage for teams in high altitude regions.

\section{Data}

Data were collected from the official NBA website. The main variables of interest
are whether or not the home team won, scoring totals for home and away teams, and
shooting percentages for home and away teams on two-pointers, three-pointers, and
free throws. These variables were very popular and frequently used in the related
literature discussed earlier. While many other measures could be used for examining
the outcome of the game and team performance, scoring seemed to be the most
important because the winner of a game is determined by who scores more points. Furthermore, 
the three types of shots are a
natural discussion point in basketball, as they're all
important and directly impact scoring totals. Free throws can often determine the winner of close
games, especially when losing teams are forced to foul the winning team to
stop the clock and hope for some missed free throws. The three
point shot has grown very prominent in basketball. The entire NBA has increased their
volume of three-point shot attempts in response to the
recent success of the Golden State Warriors and Stephen Curry. Due to this
fact, there is a growing consensus that the three-point shot is crucial because of 
the efficient scoring and floor spacing it provides offenses. 
However, two-point shots are not to be overlooked and may
actually be most important, especially for home teams, according
to \citet{Harris}. In fact, the 2020 NBA champion Los Angeles Lakers actually led the 
league in two-point shooting percentage throughout the playoffs and were bottom five
in three-point shooting percentage.

The data was collected on a game by game basis. This gave us two observations
for each variable per game played, one observation for each team(home and away).
There were 83 games played in the 2020 playoffs, giving 83 observations
for each variable in 2020 for both the home and away teams (166 observations total).
Likewise, there were 243 games played over 2017-2019, giving 243 observations of each variable
for both home and away teams over 2017-2020 (486 observations total).
 There can only be
one winner and one loser, making the outcome a binary variable, with
one indicating a win and zero indicating a loss.

Home-court advantage is the basic idea that the home team is more likely to win given 
they benefit from positive effects of a few factors we discussed earlier.
So laying a foundation of typical home-court advantage is crucial. Before
focusing on the 2017 to 2020 playoffs, we can take a quick look at home team win
percentages since 2010. Notice in Figure \ref{fig:Fig1}, the 10 years before 2020, the home team
winning percentage ranged from around 0.56 to 0.7 and never dipped below 0.5. The
2020 bubble broke this historic pattern, dipping down below 0.5. Foreshadowing
the confirmation of the expectation that the effect of home-court advantage was removed
in the 2020 playoffs.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.4\linewidth]{Fig1}
  \caption{Winning percentage of NBA home teams in the playoffs since
    2010, the green line denotes .500.}
  \label{fig:Fig1}
\end{figure}

Moving on to the main focus of the study, comparing 2020 to 2017-2019.
Figure \ref{fig:Fig2} shows the histograms of the home (green) and
away scoring (red) for 2020 vs.~2017-2019. All histograms are
fairly bell shaped, which is important for statistical tests
designed for normally distributed data. There appears to be little
difference between the 2020 and 2017-2019 for home scoring. For
away scoring, a more pronounced shift to the right in 2020 is observed
compared to 2017-2019.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.4\linewidth]{Fig2}
  \caption{Histograms of home (blue) and away (red) scoring
    for 2020 (bottom) and 2017-2019 (top).}
  \label{fig:Fig2}
\end{figure}

Our second target of inference is shooting percentage for home and away teams.
Figure \ref{fig:Fig3} shows home shooting for two-pointers, three-pointers and
free throws for 2020 (top) vs.~2017-19 (bottom). The histograms appear to be fairly
similarly distributed between 2020 and 2017-19. Likewise, Figure \ref{fig:Fig4},
shows the same percentages except for away teams. It appears that the
two-point shooting percentage for away teams has a small shift to the
right in 2020 relative to 2017-2019.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.4\linewidth]{Fig3}
  \caption{Histograms of home shooting percentages for two-
    pointers, three-pointers and free throws for 2020 (top) vs.~2017-19
    (bottom).}
  \label{fig:Fig3}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.4\linewidth]{Fig4}
  \caption{Histograms of away shooting percentages for two-pointers,
    three-pointers and free throws for 2020 (top) vs.~2017-19
    (bottom).}
  \label{fig:Fig4}  
\end{figure}

\section{Methods}

The 2020 bubble provides a new and exciting opportunity to study home-court
advantage for the NBA. Unlike college basketball, aside from a few
exhibition/preseason games, the NBA always has a home and away team. So, for
the first time in NBA history, the bubble allows NBA home and away performance
to be compared against a control/neutral field. The NBA bubble
removed many, if not all, factors impacting home-court advantage. The NBA 
bubble featured 8 seeding games for each team, then a
standard playoff format. The focus of this study was on the play during the playoff
games, since it followed the standard playoff format and could be compared
back to other playoffs. For this study, the 2020 playoffs were compared against
the three previous playoffs collectively. To control
for the changing play style of the NBA, we limit the study to 2020 vs 2017-2019
due to the faster pace of play and more common use of the three-point shot
in modern basketball. If we used data from say 10 years ago, or earlier, observed
differences may not be from effects of the NBA bubble, but rather from the effects
of drastic changes in the style of play between the seasons. However, basketball
evolves slow enough that we can reasonably assume 2017-2019 are at least very
close in pace and playing style to 2020.

Comparisons between 2020 and 2017-19 home and away teams were made on home team
winning percentage, total team scoring and two-point, three-point and free throw
shooting. Comparing the differences in these metrics for home and away teams in
2020 vs previous years will provide valuable insights to the understanding of
home-court advantage. We can see how going on the road may negatively impact
away performance and how playing at home may positively impact home performance.
If there are differences in scoring for home or away teams, the differences can
be used to show how home-court advantage affects the overall performance of home
and away teams. While testing for differences in shooting will provide added
context for how home-court advantage specifically affects performance.
Shooting percentages are not the only possible metrics affected by home-court
advantage, but they are the most obvious and likely most important one.

@@

We also conducted nonparametric tests that are distribution free to
confirm the results from the \(z\)-test. For question 1, we used
Fisher's exact test for a contingency table which summarizes the wins
and losses of the home team in the 83 games in 2020 and the 243 games
in 2017-2019. For all other eight questions, the data are the scores
or shooting percentages from the 83 games in 2020 and the 243 games in
2017-2019. We used Wilcoxan's rank-sum test.

@@

Regression was also considered, but it did not seem to add any additional 
useful insight to answering the questions presented. Regression is a popular
tool in home-court advantage studies, but as mentioned previously, this
study is very different from past studies. Rather than trying to prove the
existence of home-court advantage, like many previous studies, this paper
was more interested in generating a clearer understanding of how home-court
advantage affects each team. The most effective way to do that is hypothesis
testing to compare performance in the neutral bubble to previous years with
normal game conditions for both the home and away teams. Another potential
benefit of regression would have been the ability to control for factors
like the strength of opponent. However, we felt that the strength of team was
already well enough controlled for by the fact that we focused on the
playoffs. The playoffs only include the strongest 16 of 30 teams, then
continue to remove the less competitive teams, so that the talent disparity
between teams is much smaller than regular season play, where controlling for
opponent strength is more likely to be important.
